---
title: "01 Original Data"
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
bibliography: "../../transcriptomics-benchmarks.bib"
biblio-style: apalike
link-citations: true
output:
  html_document:
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit:
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile, encoding = encoding,
      output_file = "../../results/habib-2017-snRNAseq/01-original-data.html")})
---

```{r, include = FALSE}
#    This file is part of transcriptomics-benchmarks.
#    Copyright (C) 2019  Emir Turkes
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

knitr::opts_chunk$set(fig.width = 8.5, fig.height = 7)
```

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of the [Transcriptomics Benchmarks Project](https://github.com/eturkes/transcriptomics-benchmarks).*

In this project we apply benchmarks on the cleaned dataset derived from the analysis in [github.com/eturkes/habib-2017-snRNAseq](https://github.com/eturkes/habib-2017-snRNAseq), which uses droplet-based UMI data from @habib_massively_2017.
In this particular document we view the original dataset as well as apply some techniques without prior cleaning/dimensionality reduction.
We start by setting some global variables and loading in any required packages.

```{r}
data_str <- "habib-2017-snRNAseq"
data_str_sub <- unlist(strsplit(data_str, "-"))[1]
assets_dir <- file.path(getwd(), "..", "..", "assets", data_str)
results_dir <- file.path(getwd(), "..", "..", "results", data_str)

packages <- c("conflicted", "scater")
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))

options(stringsAsFactors = FALSE)

# Create a unique cache and results data directory for each iterated section.
if (!dir.exists(file.path(assets_dir, "cache", "01"))) {
  dir.create(file.path(assets_dir, "cache", "01"), recursive = TRUE)}
if (!dir.exists(file.path(results_dir, "data", "01"))) {
  dir.create(file.path(results_dir, "data", "01"), recursive = TRUE)}
```

As cluster labels and tSNE coordinates were released by the authors, Figure 2a and 2b were recreated in the analysis, along with their successful replication in addition to Figure 2d.
Please see the analysis for details of the processing, which was performed in close accordance to what was done in @habib_massively_2017.

```{r}
sce <- readRDS(file.path(assets_dir, "sce.rds"))
sce
```

# Habib Clusters Habib tSNE

```{r}
reducedDims(sce) <- SimpleList(
  PCA = reducedDim(sce, "PCA"),
  TSNE = data.frame(colData(sce)[ , paste0(data_str_sub, "_tsne", 1:2)]))
plotTSNE(sce, colour_by = paste0(data_str_sub, "_cluster_name"))
```

# HVG Seurat Clusters Habib tSNE

```{r}
plotTSNE(sce, colour_by = "seurat_hvg_clusters_name")
```

# HVG Seurat Clusters HVG tSNE

```{r}
reducedDims(sce) <- SimpleList(
  PCA = reducedDim(sce, "PCA"), TSNE = data.frame(colData(sce)[ , paste0("hvg_tsne", 1:2)]))
plotTSNE(sce, colour_by = paste0(data_str_sub, "_cluster_name"))
```

# Habib Clusters HVG tSNE

```{r}
plotTSNE(sce, colour_by = "seurat_hvg_clusters_name")
```

# References

This is the concluding section of the document. Here we write relevant results to disk, output the sessionInfo, and create a bibliography for works cited.

```{r}
sessionInfo()
```
