---
title: '04 HVGs - `r unlist(strsplit(getwd(), "/"))[6]`'
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
bibliography: '../../`r unlist(strsplit(getwd(), "/"))[4]`.bib'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(
    "..", "..", "results", unlist(strsplit(getwd(), "/"))[6], "04-hvg.html"
  ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of the [Transcriptomics Benchmarks Project](https://github.com/eturkes/transcriptomics-benchmarks), which aims to benchmark a wide variety of transcriptomics tools on real and simulated datasets.*

In this particular document we assess methods for extracting highly variable genes or HVGs.
The data here is derived from @`r unlist(strsplit(getwd(), "/"))[6]` and will be referenced using the name ``r unlist(strsplit(getwd(), "/"))[6]``.

```{r}
# Load in necessary boilerplate and libraries.
# --------------------------------------------

#    This file is part of transcriptomics-benchmarks.
#    Copyright (C) 2019  Emir Turkes
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

analysis_no <- 4
data_name <- unlist(strsplit(getwd(), "/"))[6] # Name of dataset.
data_name_stem <- unlist(strsplit(data_name, "-"))[1] # `data_name` up to the first hyphen.
assets_dir <- file.path(getwd(), "..", "..", "assets", data_name) # Misc binaries and temp files.
results_dir <- file.path(getwd(), "..", "..", "results", data_name)

# Unique cache and results directory for each analysis number.
if (!dir.exists(file.path(assets_dir, "cache", paste0("0", analysis_no)))) {
  dir.create(file.path(assets_dir, "cache", paste0("0", analysis_no)), recursive = TRUE)
}
if (!dir.exists(file.path(results_dir, "data", paste0("0", analysis_no)))) {
  dir.create(file.path(results_dir, "data", paste0("0", analysis_no)), recursive = TRUE)
}

packages <- c(
  "conflicted", "SingleCellExperiment", "scater", "scran", "data.table", "DT", "Seurat", "future"
)
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))
source(file.path(getwd(), "..", "utils.R"))

options(stringsAsFactors = FALSE)
plan("multiprocess") # Parallel processing for Seurat.

# Use output of previous analysis.
sce <- readRDS(file.path(results_dir, "data", paste0("0", analysis_no - 1), "sce.rds"))
```

First, let's take a look at some feature-level summary statistics.

```{r}
plotHighestExprs(sce, colour_cells_by = "batch")
```

# Technical Noise Quantification

By modeling the Poisson technical noise and blocking on batch, we get the top 10% of genes with the largest biological component values.

```{r}
set.seed(1)
dec <- modelGeneVarByPoisson(sce, block = sce$batch)
top_dec <- as.data.table(dec[order(dec$bio, decreasing = TRUE), 1:6][1:10, ])
rownames(top_dec) <- rownames(dec[order(dec$bio, decreasing = TRUE), 1:6][1:10, ])
datatable_custom(top_dec)

plotExpression(sce, features = rownames(top_dec)) +
  stat_summary(
    fun.y = median, fun.ymin = median, fun.ymax = median,
    geom = "crossbar", width = 0.3, alpha = 0.8
  )
```

```{r, fig.height = 20}
rm(top_dec)

par(mfrow = c(11, 2))
blocked_stats <- dec$per.block
for (i in colnames(blocked_stats)) {
  current <- blocked_stats[[i]]
  plot(
    current$mean, current$total, main = i, pch = 16,
    cex = 0.5, xlab = "Mean of log-expression", ylab = "Variance of log-expression")
  curve_fit <- metadata(current)
  points(curve_fit$mean, curve_fit$var, col = "red", pch = 16)
  curve(curve_fit$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)}
```

```{r}
hvg <- getTopHVGs(dec, prop = 0.1)
hvg_sce <- sce[hvg, ]
altExp(hvg_sce, "orig") <- sce
rm(blocked_stats, current, curve_fit)
dim(hvg_sce)

# Cache the results.
rds <- file.path(assets_dir, "cache", paste0("0", analysis_no), "pca.rds")
if (file.exists(rds)) {
  hvg_sce <- readRDS(rds)
} else {
  set.seed(1)
  hvg_sce <- runPCA(hvg_sce, BPPARAM = MulticoreParam())
  add_df <- data.frame(reducedDim(hvg_sce, "PCA")[ , 1:2])
  names(add_df) <- paste0("hvg_pca", seq(ncol(add_df)))
  colData(hvg_sce) <- cbind(
    colData(hvg_sce), hvg_pca1 = add_df$hvg_pca1, hvg_pca2 = add_df$hvg_pca2
  )
  saveRDS(hvg_sce, rds)
}

red_dim_plot(hvg_sce, "hvg_pca1", "hvg_pca2", paste0(data_name_stem, "_cluster_name"))
red_dim_plot(hvg_sce, "hvg_pca1", "hvg_pca2", "batch")
```

# sctransform

An advanced normalization method available in Seurat that also returns HVGs (3,000 by default) as well as regression against confounding factors.

```{r}
# Cache the results.
rds <- file.path(assets_dir, "cache", paste0("0", analysis_no), "sctransform.rds")
if (file.exists(rds)) {
  seurat <- readRDS(rds)
} else {
  seurat <- as.Seurat(sce, verbose = FALSE)
  seurat <- SCTransform(
    seurat, vars.to.regress = c("subsets_mito_percent", "subsets_ribo_percent"), verbose = FALSE
  )
  seurat <- RunPCA(seurat, verbose = FALSE)
  add_df <- data.frame(Embeddings(seurat, reduction = "pca")[ , 1:2])
  names(add_df) <- paste0("hvg_pca", seq(ncol(add_df)))
  seurat$hvg_pca1 <- add_df$hvg_pca1
  seurat$hvg_pca2 <- add_df$hvg_pca2
  rm(add_df)
  saveRDS(seurat, rds)
}
rm(sce)

red_dim_plot(seurat, "hvg_pca1", "hvg_pca2", paste0(data_name_stem, "_cluster_name"))
red_dim_plot(seurat, "hvg_pca1", "hvg_pca2", "batch")
```

# References

This is the concluding section of the document. Here we write relevant results to disk, output the `sessionInfo`, and create a bibliography for works cited.

```{r}
saveRDS(hvg_sce, file.path(results_dir, "data", paste0("0", analysis_no), "sce.rds"))
saveRDS(seurat, file.path(results_dir, "data", paste0("0", analysis_no), "seurat.rds"))
saveRDS(dec, file.path(results_dir, "data", paste0("0", analysis_no), "dec.rds"))

sessionInfo()
```
