---
title: '06 Nonlinear Reduction - `r unlist(strsplit(getwd(), "/"))[6]`'
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
bibliography: '../../`r unlist(strsplit(getwd(), "/"))[4]`.bib'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(
    "..", "..", "results", unlist(strsplit(getwd(), "/"))[6], "05-nonlinear.html"
  ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of the [Transcriptomics Benchmarks Project](https://github.com/eturkes/transcriptomics-benchmarks), which aims to benchmark a wide variety of transcriptomics tools on real and simulated datasets.*

In this particular document we assess nonlinear dimensionality reduction methods.
The data here is derived from @`r unlist(strsplit(getwd(), "/"))[6]` and will be referenced using the name ``r unlist(strsplit(getwd(), "/"))[6]``.

```{r}
# Load in necessary boilerplate and libraries.
# --------------------------------------------

#    This file is part of transcriptomics-benchmarks.
#    Copyright (C) 2019  Emir Turkes
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

analysis_no <- 6
data_name <- unlist(strsplit(getwd(), "/"))[6] # Name of dataset.
data_name_stem <- unlist(strsplit(data_name, "-"))[1] # `data_name` up to the first hyphen.
assets_dir <- file.path(getwd(), "..", "..", "assets", data_name) # Misc binaries and temp files.
results_dir <- file.path(getwd(), "..", "..", "results", data_name)

# Unique cache and results directory for each analysis number.
if (!dir.exists(file.path(assets_dir, "cache", paste0("0", analysis_no)))) {
  dir.create(file.path(assets_dir, "cache", paste0("0", analysis_no)), recursive = TRUE)
}
if (!dir.exists(file.path(results_dir, "data", paste0("0", analysis_no)))) {
  dir.create(file.path(results_dir, "data", paste0("0", analysis_no)), recursive = TRUE)
}

packages <- c(
  "conflicted", "SingleCellExperiment", "scater", "magrittr", "dplyr", "ggrepel", "Seurat", "future"
)
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))
source(file.path(getwd(), "..", "utils.R"))

options(stringsAsFactors = FALSE)
plan("multiprocess") # Parallel processing for Seurat.

# Use output of previous analysis.
sce <- readRDS(file.path(results_dir, "data", paste0("0", analysis_no - 1), "sce.rds"))
seurat <- readRDS(file.path(results_dir, "data", paste0("0", analysis_no - 1), "seurat.rds"))
```

# SCE

```{r}
# Cache the results.
rds <- file.path(assets_dir, "cache", paste0("0", analysis_no), "nonlinear_sce.rds")
if (file.exists(rds)) {
  sce <- readRDS(rds)
} else {
  set.seed(1)
  sce <- runTSNE(sce, BPPARAM = MulticoreParam())
  set.seed(1)
  sce <- runUMAP(sce, min_dist = 0.3, BPPARAM = MulticoreParam()) # Match Seurat's defaults.
  
  add_df <- data.frame(reducedDim(sce, "TSNE"))
  names(add_df) <- paste0("tsne", seq(ncol(add_df)))
  colData(sce) <- cbind(colData(sce), tsne1 = add_df$tsne1, tsne2 = add_df$tsne2)
  
  add_df <- data.frame(reducedDim(sce, "UMAP"))
  names(add_df) <- paste0("umap", seq(ncol(add_df)))
  colData(sce) <- cbind(colData(sce), umap1 = add_df$umap1, umap2 = add_df$umap2)
  
  saveRDS(sce, rds)
}
```

## tSNE

```{r}
red_dim_plot(sce, "tsne1", "tsne2", paste0(data_name_stem, "_cluster_name"), "cat")
red_dim_plot(sce, "tsne1", "tsne2", "batch")
```

## UMAP

```{r}
red_dim_plot(sce, "umap1", "umap2", paste0(data_name_stem, "_cluster_name"), "cat")
red_dim_plot(sce, "umap1", "umap2", "batch")
```

# Seurat

```{r}
# Cache the results.
rds <- file.path(assets_dir, "cache", paste0("0", analysis_no), "nonlinear_seurat.rds")
if (file.exists(rds)) {
  sce <- readRDS(rds)
} else {
  seurat <- RunTSNE(seurat, dims = 1:13, verbose = FALSE)
  seurat <- RunUMAP(seurat, dims = 1:13, verbose = FALSE)
  
  add_df <- data.frame(Embeddings(seurat, reduction = "tsne"))
  names(add_df) <- paste0("tsne", seq(ncol(add_df)))
  seurat$tsne1 <- add_df$tsne1
  seurat$tsne2 <- add_df$tsne2
  
  add_df <- data.frame(Embeddings(seurat, reduction = "umap"))
  names(add_df) <- paste0("umap", seq(ncol(add_df)))
  seurat$umap1 <- add_df$umap1
  seurat$umap2 <- add_df$umap2
  
  rm(add_df)
  saveRDS(seurat, rds)
}
```

## tSNE

```{r}
red_dim_plot(seurat, "tsne1", "tsne2", paste0(data_name_stem, "_cluster_name"), "cat")
red_dim_plot(seurat, "tsne1", "tsne2", "batch")
```

## UMAP

```{r}
red_dim_plot(seurat, "umap1", "umap2", paste0(data_name_stem, "_cluster_name"), "cat")
red_dim_plot(seurat, "umap1", "umap2", "batch")
```

# References

This is the concluding section of the document. Here we write relevant results to disk, output the `sessionInfo`, and create a bibliography for works cited.

```{r}
saveRDS(sce, file.path(results_dir, "data", paste0("0", analysis_no), "sce.rds"))
saveRDS(seurat, file.path(results_dir, "data", paste0("0", analysis_no), "seurat.rds"))

sessionInfo()
```
