---
title: "02 PCA"
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
bibliography: "../../transcriptomics-benchmarks.bib"
biblio-style: apalike
link-citations: true
output:
  html_document:
    code_folding: show
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
knit:
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile, encoding = encoding,
      output_file = "../../results/lake-2016-snRNAseq/02-pca.html")})
---

```{r, include = FALSE}
#    This file is part of transcriptomics-benchmarks.
#    Copyright (C) 2019  Emir Turkes
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

knitr::opts_chunk$set(fig.width = 8.5, fig.height = 7)
```

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.90em; padding-left: 35px; text-indent: 0;}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of the [Transcriptomics Benchmarks Project](https://github.com/eturkes/transcriptomics-benchmarks).*

This report assesses the performance of several PCA methods.
Please see @tsuyuzaki_benchmarking_2019 for a good review of methods used in scRNAseq.
We start by setting some global variables and loading in any required packages.

```{r}
assets_dir <- file.path(getwd(), "..", "..", "assets/lake-2016-snRNAseq")
results_dir <- file.path(getwd(), "..", "..", "results/lake-2016-snRNAseq")

packages <- c(
  "conflicted", "SingleCellExperiment", "magrittr", "dplyr", "ggplot2", "ggrepel", "S4Vectors",
  "SummarizedExperiment", "DropletUtils", "scran", "BiocSingular", "scater", "Rtsne", "svd",
  "SC3", "DT", "data.table", "Seurat", "uwot", "viridis", "rsvd")
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))

conflict_prefer("which", "BiocGenerics")
options(stringsAsFactors = FALSE)

# Create a unique cache and results data directory for each iterated section.
if (!dir.exists(file.path(assets_dir, "cache", "02"))) {
  dir.create(file.path(assets_dir, "cache", "02"), recursive = TRUE)}
if (!dir.exists(file.path(results_dir, "data", "02"))) {
  dir.create(file.path(results_dir, "data", "02"), recursive = TRUE)}

# ggplot2 function providing custom aesthetics and automatic placement of categorical labels.
# For continuous data, a colorbar is implemented.
dim_red_plot <- function(data, x, y, col, type) {
  gg <- ggplot(data, aes_string(x = x, y = y, color = col)) +
    geom_point(alpha = 0.35, stroke = 0.05, shape = 21, aes_string(fill = col)) +
    theme_classic() +
    theme(
      legend.position = "right", plot.title = element_text(hjust = 0.5),
      legend.title = element_blank()) +
    guides(color = guide_legend(override.aes = list(alpha = 1)))
    if (type == "cat") {
      gg <- gg + geom_label_repel(data = label_df2, aes(label = label), show.legend = FALSE)
    } else if (type == "cont") {
      gg <- ggplot(data, aes_string(x = x, y = y)) +
        geom_point(alpha = 0.35, stroke = 0.05, aes_string(color = col)) +
        theme_classic() +
        theme(
          legend.position = "right", plot.title = element_text(hjust = 0.5),
          legend.title = element_blank()) +
        scale_colour_viridis()}
  gg}

# Adds download buttons.
datatable_custom <- function(dt) {
  datatable(
    dt,
    extensions = "Buttons", options = list(dom = "Blfrtip", buttons = list(
      "copy", "print",
      list(extend = "collection", buttons = c("csv", "excel", "pdf"), text = "Download"))))}
```

We use the SCE object from the analysis in [github.com/eturkes/lake-2016-snRNAseq](https://github.com/eturkes/lake-2016-snRNAseq).

```{r}
sce <- readRDS(file.path(assets_dir, "sce_orig.rds"))
sce
```

# prcomp

```{r}
# Knitr cache turned off due to error:
# Error in lazyLoadDBinsertValue(data, datafile, ascii, compress, envhook) :
# long vectors not supported yet: connections.c:5984
rds <- file.path(assets_dir, "cache", "02", "prcomp.rds")
if (file.exists(rds)) {
  pca <- readRDS(rds)
} else {
  set.seed(1)
  pca <- runPCA(sce, ncomponents = 50)
  saveRDS(pca, rds)}
```

```{r}
par(mfrow = c(1, 1))
plot(
  log10(attr(reducedDim(pca, "PCA"), "percentVar")), xlab = "PC",
  ylab = "log10(Proportion of Variance Explained)", pch = 20,
  cex = 0.6, col = rgb(0.8, 0.2, 0.2, 0.5))
abline(v = c(10, 20, 30, 40, 50), lty = 2, col = "red")
```

## 10 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "prcomp10_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(reducedDim(pca, "PCA")[ , 1:10], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = reducedDim(pca, "PCA"), UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("prcomp10_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), prcomp10_umap1 = add_df$prcomp10_umap1,
  prcomp10_umap2 = add_df$prcomp10_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    prcomp10_umap1 = median(prcomp10_umap1),
    prcomp10_umap2 = median(prcomp10_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "prcomp10_umap1", y = "prcomp10_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters prcomp 10 UMAP")
```

## 20 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "prcomp20_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(reducedDim(pca, "PCA")[ , 1:20], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = reducedDim(pca, "PCA"), UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("prcomp20_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), prcomp20_umap1 = add_df$prcomp20_umap1,
  prcomp20_umap2 = add_df$prcomp20_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    prcomp20_umap1 = median(prcomp20_umap1),
    prcomp20_umap2 = median(prcomp20_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "prcomp20_umap1", y = "prcomp20_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters prcomp 20 UMAP")
```

## 30 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "prcomp30_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(reducedDim(pca, "PCA")[ , 1:30], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = reducedDim(pca, "PCA"), UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("prcomp30_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), prcomp30_umap1 = add_df$prcomp30_umap1,
  prcomp30_umap2 = add_df$prcomp30_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    prcomp30_umap1 = median(prcomp30_umap1),
    prcomp30_umap2 = median(prcomp30_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "prcomp30_umap1", y = "prcomp30_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters prcomp 30 UMAP")
```

## 40 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "prcomp40_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(reducedDim(pca, "PCA")[ , 1:40], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = reducedDim(pca, "PCA"), UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("prcomp40_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), prcomp40_umap1 = add_df$prcomp40_umap1,
  prcomp40_umap2 = add_df$prcomp40_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    prcomp40_umap1 = median(prcomp40_umap1),
    prcomp40_umap2 = median(prcomp40_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "prcomp40_umap1", y = "prcomp40_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters prcomp 40 UMAP")
```

## 50 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "prcomp50_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(reducedDim(pca, "PCA")[ , 1:50], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = reducedDim(pca, "PCA"), UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("prcomp50_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), prcomp50_umap1 = add_df$prcomp50_umap1,
  prcomp50_umap2 = add_df$prcomp50_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    prcomp50_umap1 = median(prcomp50_umap1),
    prcomp50_umap2 = median(prcomp50_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "prcomp50_umap1", y = "prcomp50_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters prcomp 50 UMAP")
```

# rpca

Randomized principal component analysis as implemented by the `rpca` function in the `rsvd` package.

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca.rds")
if (file.exists(rds)) {
  pca <- readRDS(rds)
} else {
  set.seed(1)
  pca <- rpca(scale(t(as.matrix(normcounts(sce)))), k = 50)
  saveRDS(pca, rds)}
```

```{r}
par(mfrow = c(1, 1))
plot(
  log10(pca$sdev), xlab = "PC", ylab = "log10(Proportion of Variance Explained)",
  pch = 20, cex = 0.6, col = rgb(0.8, 0.2, 0.2, 0.5))
abline(v = c(10, 20, 30, 40, 50), lty = 2, col = "red")
```

## 10 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca10_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca$x[ , 1:10], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca$x, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("rpca10_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), rpca10_umap1 = add_df$rpca10_umap1,
  rpca10_umap2 = add_df$rpca10_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    rpca10_umap1 = median(rpca10_umap1), rpca10_umap2 = median(rpca10_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "rpca10_umap1", y = "rpca10_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters rpca 10 UMAP")
```

## 20 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca20_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca$x[ , 1:20], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca$x, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("rpca20_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), rpca20_umap1 = add_df$rpca20_umap1,
  rpca20_umap2 = add_df$rpca20_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    rpca20_umap1 = median(rpca20_umap1), rpca20_umap2 = median(rpca20_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "rpca20_umap1", y = "rpca20_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters rpca 20 UMAP")
```

## 30 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca30_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca$x[ , 1:30], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca$x, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("rpca30_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), rpca30_umap1 = add_df$rpca30_umap1,
  rpca30_umap2 = add_df$rpca30_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    rpca30_umap1 = median(rpca30_umap1), rpca30_umap2 = median(rpca30_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "rpca30_umap1", y = "rpca30_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters rpca 30 UMAP")
```

## 40 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca40_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca$x[ , 1:40], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca$x, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("rpca40_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), rpca40_umap1 = add_df$rpca40_umap1,
  rpca40_umap2 = add_df$rpca40_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    rpca40_umap1 = median(rpca40_umap1), rpca40_umap2 = median(rpca40_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "rpca40_umap1", y = "rpca40_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters rpca 40 UMAP")
```

## 50 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "rpca50_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca$x[ , 1:50], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca$x, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("rpca50_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), rpca50_umap1 = add_df$rpca50_umap1,
  rpca50_umap2 = add_df$rpca50_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(
    rpca50_umap1 = median(rpca50_umap1), rpca50_umap2 = median(rpca50_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "rpca50_umap1", y = "rpca50_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters rpca 50 UMAP")
```

# svd

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd.rds")
if (file.exists(rds)) {
  svd <- readRDS(rds)
} else {
  set.seed(1)
  svd <- svd(scale(t(as.matrix(normcounts(sce)))), nv = 50) # TODO: Correctly calculate for 50 PCs.
  saveRDS(svd, rds)}
```

```{r}
pca <- t(svd$d * t(svd$u))
par(mfrow = c(1, 1))
plot(
  log10(svd$d), xlab = "PC", ylab = "log10(Proportion of Variance Explained)",
  pch = 20, cex = 0.6, col = rgb(0.8, 0.2, 0.2, 0.5))
abline(v = c(10, 20, 30, 40, 50), lty = 2, col = "red")
```

## 10 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd10_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:10], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("svd10_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), svd10_umap1 = add_df$svd10_umap1,
  svd10_umap2 = add_df$svd10_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(svd10_umap1 = median(svd10_umap1), svd10_umap2 = median(svd10_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "svd10_umap1", y = "svd10_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters SVD 10 UMAP")
```

## 20 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd20_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:20], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("svd20_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), svd20_umap1 = add_df$svd20_umap1,
  svd20_umap2 = add_df$svd20_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(svd20_umap1 = median(svd20_umap1), svd20_umap2 = median(svd20_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "svd20_umap1", y = "svd20_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters SVD 20 UMAP")
```

## 30 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd30_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:30], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("svd30_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), svd30_umap1 = add_df$svd30_umap1,
  svd30_umap2 = add_df$svd30_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(svd30_umap1 = median(svd30_umap1), svd30_umap2 = median(svd30_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "svd30_umap1", y = "svd30_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters SVD 30 UMAP")
```

## 40 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd40_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:40], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("svd40_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), svd40_umap1 = add_df$svd40_umap1,
  svd40_umap2 = add_df$svd40_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(svd40_umap1 = median(svd40_umap1), svd40_umap2 = median(svd40_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "svd40_umap1", y = "svd40_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters SVD 40 UMAP")
```

## 50 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "svd50_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:50], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("svd50_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), svd50_umap1 = add_df$svd50_umap1,
  svd50_umap2 = add_df$svd50_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(svd50_umap1 = median(svd50_umap1), svd50_umap2 = median(svd50_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "svd50_umap1", y = "svd50_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters SVD 50 UMAP")
```

# propack.svd

We use PROPACK SVD to generate PCs, similar to the implementation used in GiniClust2 [@tsoucas_giniclust2_2018].

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk.rds")
if (file.exists(rds)) {
  ppk <- readRDS(rds)
} else {
  set.seed(1)
  ppk <- propack.svd(scale(t(as.matrix(normcounts(sce)))), neig = 50)
  saveRDS(ppk, rds)}
```

```{r}
pca <- t(ppk$d * t(ppk$u))
par(mfrow = c(1, 1))
plot(
  log10(ppk$d), xlab = "PC", ylab = "log10(Proportion of Variance Explained)",
  pch = 20, cex = 0.6, col = rgb(0.8, 0.2, 0.2, 0.5))
abline(v = c(10, 20, 30, 40, 50), lty = 2, col = "red")
```

## 10 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk10_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:10], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("ppk10_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), ppk10_umap1 = add_df$ppk10_umap1,
  ppk10_umap2 = add_df$ppk10_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(ppk10_umap1 = median(ppk10_umap1), ppk10_umap2 = median(ppk10_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "ppk10_umap1", y = "ppk10_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters PPK 10 UMAP")
```

## 20 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk20_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:20], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("ppk20_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), ppk20_umap1 = add_df$ppk20_umap1,
  ppk20_umap2 = add_df$ppk20_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(ppk20_umap1 = median(ppk20_umap1), ppk20_umap2 = median(ppk20_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "ppk20_umap1", y = "ppk20_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters PPK 20 UMAP")
```

## 30 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk30_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:30], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("ppk30_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), ppk30_umap1 = add_df$ppk30_umap1,
  ppk30_umap2 = add_df$ppk30_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(ppk30_umap1 = median(ppk30_umap1), ppk30_umap2 = median(ppk30_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "ppk30_umap1", y = "ppk30_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters PPK 30 UMAP")
```

## 40 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk40_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:40], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("ppk40_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), ppk40_umap1 = add_df$ppk40_umap1,
  ppk40_umap2 = add_df$ppk40_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(ppk40_umap1 = median(ppk40_umap1), ppk40_umap2 = median(ppk40_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "ppk40_umap1", y = "ppk40_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters PPK 40 UMAP")
```

## 50 PCs

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "ppk50_umap.rds")
if (file.exists(rds)) {
  umap <- readRDS(rds)
} else {
  set.seed(1)
  umap <- umap(pca[ , 1:50], min_dist = 0.5)
  saveRDS(umap, rds)}
```

```{r}
reducedDims(sce) <- SimpleList(PCA = pca, UMAP = umap)
add_df <- data.frame(reducedDim(sce, "UMAP"))
names(add_df) <- paste0("ppk50_umap", seq(ncol(add_df)))
colData(sce) <- cbind(
  colData(sce), ppk50_umap1 = add_df$ppk50_umap1,
  ppk50_umap2 = add_df$ppk50_umap2)

# Set up data frame for ggplot2.
gg_df <- data.frame(colData(sce)[ , c("lake_cluster_name", "region")], add_df)
rownames(gg_df) <- NULL
gg_df$lake_cluster_name <- factor(gg_df$lake_cluster_name)

# Setup for automatic placement of cluster labels.
label_df <- data.frame(
  lake_cluster_name = levels(gg_df$lake_cluster_name),
  label = levels(gg_df$lake_cluster_name))
label_df2 <- gg_df %>%
  group_by(lake_cluster_name) %>%
  summarize(ppk50_umap1 = median(ppk50_umap1), ppk50_umap2 = median(ppk50_umap2)) %>%
  left_join(label_df)

dim_red_plot(
  data = gg_df, x = "ppk50_umap1", y = "ppk50_umap2",
  col = "lake_cluster_name", type = "cat") +
  xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("Lake Clusters PPK 50 UMAP")
```

# Seurat

```{r, cache = TRUE}
rds <- file.path(assets_dir, "cache", "02", "seurat.rds")
if (file.exists(rds)) {
  seurat <- readRDS(rds)
} else {
  seurat <- as.Seurat(sce, counts = "logcounts", data = NULL)
  seurat <- FindVariableFeatures(seurat, nfeatures = dim(sce)[1])
  seurat <- ScaleData(seurat)
  seurat <- RunPCA(seurat)
  seurat <- JackStraw(seurat, dims = 50)
  seurat <- ScoreJackStraw(seurat, dims = 1:50)
  saveRDS(seurat, rds)}
```

```{r}
JackStrawPlot(seurat, dims = 1:50)
```

# References

This is the concluding section of the document. Here we write relevant results to disk, output the sessionInfo, and create a bibliography for works cited.

```{r}
sessionInfo()
```
